---
title: "687 Project"
output: html_document
---

```{r}
# Team Members: Nakul Rattan, Himanshu Mangal, Aarti Mehra, Nandita Pathardikar
```


## Analysis of HMO Health Data

# The first step is to import the CSV into a dataframe for analysis and training our model. We will also library the tidyverse package.

```{r}
library(tidyverse)

datafile <- "https://intro-datascience.s3.us-east-2.amazonaws.com/HMO_data.csv"
data <- read_csv(datafile)

```

# We can now see what the data is like. 

```{r}
str(data)
```
# Checking for null values. 

```{r}
#sum(is.na(data$X))
#sum(is.na(data$age))
#sum(is.na(data$bmi))
#sum(is.na(data$children))
#sum(is.na(data$smoker))
#sum(is.na(data$location))
#sum(is.na(data$location_type))
#sum(is.na(data$education_level))
#sum(is.na(data$yearly_physical))
#sum(is.na(data$exercise))
#sum(is.na(data$married))
#sum(is.na(data$hypertension))
#sum(is.na(data$gender))
#sum(is.na(data$cost))
```

# Interpolate the null values. 

```{r}
library(imputeTS)
data$bmi <- na_interpolation(data$bmi)
data$hypertension <- na_interpolation(data$hypertension)
```

# Adding the *expensive* column in the dataframe. 

```{r}
data = data %>% 
  mutate(expensive = if_else(cost<=median(cost)*2,FALSE, TRUE))

# Define empty columns
empty_cols <- c('ageCategory')

# Add empty columns
data[ , empty_cols] <- NA

data$ageCategory <- as.factor(ifelse(data$age<18, 'children',
                     ifelse(data$age>=18 & data$age<=34, 'young_adults',
                     ifelse(data$age>=35 & data$age<=54, 'middle_aged-adults', 'older_adults'))))
#summary(data)
```
# Let's view the data 

```{r}
#view(data)
```

# We can now run a linear regression model on the dataframe. This would also help us identify significant predictors. 

```{r}
set.seed(111)
model <- lm(expensive~.-X-cost,data=data)
summary(model)
```
# Dividing the group into expensive and not expensive

```{r}
dfExpensive <- data %>% filter(expensive==TRUE)
#dfExpensive

dfNotExpensive <- data %>% filter(expensive==FALSE)
#dfNotExpensive
```

# Checking the percentage of smokers in the expensive group

```{r}
smokersExpensive <- sum(if_else(dfExpensive$smoker=="yes",1,0))
#1063 smokers out of 1792 people in the expensive set
ratioSmokersExpensive <- smokersExpensive/nrow(dfExpensive) 

```

# Visualizing the above data in a pie chart 

```{r}
slices <- c(ratioSmokersExpensive*100, (1-ratioSmokersExpensive)*100)
lbls <- c("Smokers","Non-Smokers")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
pie(slices,labels = lbls, col=rainbow(length(lbls)),
    main="Pie Chart of Smokers vs Non-Smokers: Expensive bracket") 

```
# Checking the percentage of smokers in the not expensive group

```{r}
smokersNotExpensive <- sum(if_else(dfNotExpensive$smoker=="yes",1,0))
ratioSmokersNotExpensive <- smokersNotExpensive/nrow(dfNotExpensive)

#ratioSmokersNotExpensive

slices <- c(ratioSmokersNotExpensive*100, (1-ratioSmokersNotExpensive)*100)
lbls <- c("Smokers","Non-Smokers")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
pie(slices,labels = lbls, col=rainbow(length(lbls)),
    main="Pie Chart of Smokers vs Non-Smokers: Not Expensive bracket") 

# 7% of people in the not expensive group are smokers.

```
#Breakdown of smokers by age

``` {r}
smokersByAge <- dfExpensive %>% select(smoker,ageCategory) %>% group_by(ageCategory,smoker) %>%summarise(total_count=n(),.groups = 'drop')

ggplot(smokersByAge,aes(x=ageCategory,y=total_count,fill=smoker)) + geom_bar(stat="identity",color="black") + theme_minimal() + xlab("Age Category") + ylab("Count") + ggtitle("Breakdown of Smokers by Age: Expensive") + scale_x_discrete(limits=c("young_adults", "middle_aged-adults", "older_adults"))
```
```{r}
smokersByAge <- dfNotExpensive %>% select(smoker,ageCategory) %>% group_by(ageCategory,smoker) %>%summarise(total_count=n(),.groups = 'drop')

ggplot(smokersByAge,aes(x=ageCategory,y=total_count,fill=smoker)) + geom_bar(stat="identity",color="black") + theme_minimal() + xlab("Age Category") + ylab("Count") + ggtitle("Breakdown of Smokers by Age: Not Expensive") + scale_x_discrete(limits=c("young_adults", "middle_aged-adults", "older_adults"))

```

```{r}
#adding violin-plot of bmi wrt gender
library(ggplot2)
vioPlot <- ggplot(data, aes(x=gender, y=bmi, fill=expensive)) + 
  geom_violin(width=1,trim=FALSE) + theme_minimal() + coord_flip() + xlab("Gender") + ylab("Body Mass Index(BMI)") +
  ggtitle("Relation between BMI and Gender for health expense")

vioPlot

# From the visualization, we can see that expensive males and females have high bmi as compared to their non-expensive counterparts 
```


```{r}
# histogram plot of age for expensive data set
histPlot <- 
  ggplot(dfExpensive,aes(x=age)) +
    geom_histogram( color="#e9ecef", position = 'identity', bins=40) + theme_minimal() + xlab("Age") +  ggtitle("Breakdown by Age for Expensive treatements")

histPlot

# from the plot, expensive individuals are mostly the ones with age less than 20, age greater than 60, and also with age around 45-50.
```

```{r}
# Scatter plot of age vs bmi based on hypertension

library(ggplot2)
scatterPlot <- ggplot(dfExpensive, aes(x=age, y=bmi, color=hypertension)) + geom_point() + 
  ylab("BMI") + xlab("Age Range") + ggtitle("Age vs BMI based on Hypertension")
scatterPlot

# from the plot, it can be seen that indviduals with age greater than 60 and a high bmi are expensive, though their levels of hypertension differ, most individuals do not suffer from hypertension
```
# Map of Average Age
```{r}
library(maps); library(ggmap); library(mapproj)
```

```{r}
#unique(data$location)
us <- map_data("state")      
data$location <- tolower(data$location)

dfSimple <- data %>% group_by(location) %>% summarise(meanAge=mean(age))

dfMerged <- merge(dfSimple, us, by.x="location",by.y = "region")
dfMerged <- dfMerged %>% arrange(order)
map <- ggplot(dfMerged)          
map <- map + aes(x=long, y=lat, group=group,label=location,fill=meanAge) +  geom_polygon(color="black")
map <- map + expand_limits(x=dfMerged$long, y=dfMerged$lat)
map <- map + coord_map() + ggtitle("Average Age by State")
map

```

```{r}
# jitter plot of age vs cost with excercise

library(ggplot2)
jitterPlot <- ggplot(dfExpensive, aes(x=age, y= cost, color= exercise)) +
        geom_jitter() + xlab("Age") + ylab("Cost") + ggtitle("Age vs Cost with exercise factor")

jitterPlot

# from the plot, we can infer that most individuals pay between $10000 and $15000, with most of them being over the age of 45 and also, there are few individuals with active workout routine.
```

# Finding the correlation between activity and cost

```{r}
exerciseExpensive <- sum(if_else(dfExpensive$exercise=="Active",1,0))

ratioExerciseExpensive <- exerciseExpensive/nrow(dfExpensive)
# 13.2% people that are active are in the expensive group

slices <- c(ratioExerciseExpensive*100, (1-ratioExerciseExpensive)*100)
lbls <- c("Active","Not Active")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
pie(slices,labels = lbls,
    main="Pie Chart of Active vs Not Active: Expensive bracket") 
```

``` {r}
exerciseNotExpensive <- sum(if_else(dfNotExpensive$exercise=="Active",1,0))

ratioExerciseNotExpensive <- exerciseNotExpensive/nrow(dfNotExpensive)
# 28.5% of people who are active are in the not expensive group.

slices <- c(ratioExerciseNotExpensive*100, (1-ratioExerciseNotExpensive)*100)
lbls <- c("Active","Not Active")
pct <- round(slices/sum(slices)*100)
lbls <- paste(lbls, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels
pie(slices,labels = lbls,
    main="Pie Chart of Active vs Not Active: Not Expensive bracket")
```


## Predictive Models

```{r}

library(rpart)
library(e1071)
library(kernlab)
library(caret)

```

# Reading in the test data 

```{r}
#testData <- read_csv("D:\\Downloads\\HMO_TEST_data_sample.csv")
#testDataPrediction <- read_csv("D:\\Downloads\\HMO_TEST_data_sample_solution.csv")
#TEST_PRED = as.factor(testDataPrediction$expensive)
```

# SVM Model 

```{r}
data$expensive <- as.factor(data$expensive)

trainList <- createDataPartition(y=data$expensive,p=0.7,list=FALSE)
trainSet <- data[trainList,]
testSet <- data[-trainList,]

svmModel <- train(expensive~age+bmi+children+smoker+exercise+hypertension,data=trainSet,method="svmRadial",trControl=trainControl(method="none"),preProcess=c("center","scale"))
#svmModel

predictValues <- predict(svmModel,testSet)

confusionMatrix(predictValues, testSet$expensive)

```

# KSVM

```{r}
svm.model<-ksvm(data=trainSet,expensive~age+bmi+children+smoker+exercise+hypertension)
svm.model
predictValues <- predict(svm.model,testSet)
confusionMatrix(predictValues,testSet$expensive)
```

# SVM model with KFold #88.3% Accuracy, 97% sensitivity

```{r}
trctrl <- trainControl(method="repeatedcv",number=10)
svm.model.kfold <- train(expensive~age+bmi+children+smoker+exercise+hypertension,data=trainSet,method="svmRadial",trControl=trctrl,preProcess=c("center","scale"))
svm.model.kfold
predictValues <- predict(svm.model.kfold,testSet)
confusionMatrix(predictValues,testSet$expensive)
```

#Model with rpart - 88.92% accuracy, 97% sensitivity
```{r}
library(rpart.plot)
```

```{r}
model.rpart <- train(expensive~age+bmi+children+smoker+exercise+hypertension,method="rpart",data=trainSet,trControl=trctrl,tuneLength=50)
predictValues <- predict(model.rpart,testSet)
confusionMatrix(predictValues,testSet$expensive)

#model.rpart

rpart.plot(model.rpart$finalModel)
```

